---
title: "Reproducible Research: Peer Assessment 1"
author: "Baldvin Einarsson"
output: 
  html_document:
    keep_md: true
---

We start with some initial setup, which includes loading packages and setting global `knitr` options:

```{r setup,echo=TRUE, warning=FALSE}

library(ggplot2)
library(lubridate)
library(knitr)

opts_knit$set(echo=TRUE, comment="")

```


## Loading and preprocessing the data

The GitHub fork contains a zip file of the data, which we have unzipped. 

```{r readData}

fitData <- read.csv(file="activity.csv", header = TRUE, stringsAsFactors = FALSE)

```

Because the last question involves the days of the week, we add that information to the data frame. Also, we add a human readable form of the time of day (format hh:mm).

```{r addWeekdays}

# Both wday and ymd are from the lubridate package
fitData$weekDay <- wday(ymd(fitData$date),label = TRUE)


```

```{r addTimeOfDay,echo=FALSE,eval=FALSE}

# Figure out how many leading zeros to pad with:
leadingZeros <- sapply(FUN=paste0, 
                       X = mapply(FUN = rep, 
                                  times = 4-nchar(fitData$interval), 
                                  x = "0"), 
                       collapse='')


```


## What is mean total number of steps taken per day?

Here, we group the data by the date, and sum up the number of steps for each day. Note that we have removed the missing values.

```{r stepsPerDay}

# Group the data by day and sum up the steps, removing NA values:
stepsPerDay <- tapply(X = fitData$steps, 
                      INDEX = fitData$date, 
                      FUN = sum, 
                      na.rm=TRUE)

```

Next, we create a histogram of these values:

```{r histSteps}

ggplot(data=as.data.frame(stepsPerDay), aes(stepsPerDay)) + 
  geom_histogram(col="red", fill="blue", alpha = .8, bins = 42) + 
  labs(title="Histogram of steps taken per day") +
  labs(x="Steps", y="Count")

```

Finally, we calculate the mean and median steps per day:

```{r meanMedianSteps, results='asis'}

#NB the mean is actually a character vector after the "rounding"
meanStepsPerDay <- round(mean(stepsPerDay),digits=2)
medianStepsPerDay <- median(stepsPerDay)

```

The mean steps per day is `r meanStepsPerDay` steps/day, and the median is `r medianStepsPerDay` steps/day.

## What is the average daily activity pattern?

Similar to the previous step, we group the data by the interval, and average the number of steps for each interval (after remoing the missing values). Then, we create a data frame for plotting:

```{r AveActivity}

# Group the data by interval and average the steps, removing NA values:
aveStepsPerInterval <-  tapply(X = fitData$steps,
                               INDEX = fitData$interval, 
                               FUN = mean, 
                               na.rm=TRUE)

activity <- 
  data.frame(interval = as.numeric(names(aveStepsPerInterval)),
             aveSteps = aveStepsPerInterval,
             stringsAsFactors = FALSE
  )

```

Then, we create a time series plot of the average steps per time interval:

```{r timeSeries}

ggplot(data = activity, aes(x = interval, y = aveSteps)) + 
  geom_line() +
  labs(title="Daily activity") +
  labs(x="Interval", y="Average Steps")

```

From this time series, it appears that the most active time of the day is between 8 and 9 am. This probably corresponds to the morning commute.

Finally, we find the most active interval (in average number of step):

```{r mostActive}

mostActiveTime <- activity$interval[which.max(activity$aveSteps)]

```

The value of the `mostActiveTime` is `r mostActiveTime`.

## Imputing missing values

The total number of missing values, coded as `NA` can be found by
```{r numNA, comment=''}
numMissing <- sum(is.na(fitData$steps))
```
which has value `r numMissing`.

It also turns out that there are a few days with no recorded values. We can count those, by using the object `stepsPerDay` found above. It contains the total number of steps each day, with some having zero steps. We interpret that as the user having not worn the activity monitor that day, and we find the total number of such days in the following manner:

```{r numDaysMissing,comment=''}
sum(stepsPerDay<1)
```

> Note that we are using the fact that `sum(NA,na.rm=TRUE)` equals zero.

To impute the missing values, we generate Poisson random variables which have mean $\lambda$ equal to the average of the corresponding time interval, respectively. The average steps per interval are found in the object `aveStepsPerInterval`, created above.

```{r impute}

aveStepsAtMissingVal <- 
  activity$aveSteps[match(fitData$interval[is.na(fitData$steps)],
                          activity$interval)]

set.seed(42)
imputedValues <- mapply(rpois, n=1, lambda= aveStepsAtMissingVal)

# First, we copy the fitData data.frame:
fitDataNoNA <- fitData
# Then, we fill in the missing values:
fitDataNoNA$steps[is.na(fitData$steps)] <- imputedValues

```

For the newly created data frame `fitDataNoNA` we group the data by the date, and sum up the number of steps for each day.

```{r stepsPerDayNoNA}

# Group the data by day and sum up the steps:
stepsPerDayNoNA <- tapply(X = fitDataNoNA$steps, 
                      INDEX = fitDataNoNA$date, 
                      FUN = sum)
# NB: We did not need to remove any NA values

```

Next, we create a histogram of these values:

```{r histStepsNoNA}

ggplot(data=as.data.frame(stepsPerDayNoNA), aes(stepsPerDayNoNA)) + 
  geom_histogram(col="red", fill="blue", alpha = .8, bins = 42) + 
  labs(title="Histogram of steps taken per day (Imputed)") +
  labs(x="Steps", y="Count")

```

Finally, we calculate the mean and median steps per day for the imputed data:

```{r meanMedianStepsNoNA, results='asis'}

#NB the mean is actually a character vector after the "rounding"
meanStepsPerDayNoNA <- round(mean(stepsPerDayNoNA),digits=2)
medianStepsPerDayNoNA <- median(stepsPerDayNoNA)

```

For the imputed data, the mean steps per day is `r meanStepsPerDayNoNA` steps/day, and the median is `r medianStepsPerDayNoNA` steps/day. We note that these values are slightly higher as those for the raw data. This is reasonable because missing values do not contribute to the computations.

## Are there differences in activity patterns between weekdays and weekends?

Using the newly created data frame, with missing 
